High-Fidelity Counterfactual Explanations through Conformal Prediction

### 2023-04-25 notebooks/mnist.qmd
MNIST

```{julia}
# Data:
n_obs = 1000
counterfactual_data = load_mnist(n_obs)
counterfactual_data.X = pre_process.(counterfactual_data.X)
X, y = CounterfactualExplanations.DataPreprocessing.unpack_data(counterfactual_data)
X = table(permutedims(X))
labels = counterfactual_data.output_encoder.labels
input_dim, n_obs = size(counterfactual_data.X)
n_digits = Int(sqrt(input_dim))
output_dim = length(unique(labels))
```

@@

```{julia}
for (mod_name, mod) in model_dict

    # Plot:
    if mod.model.model isa JointEnergyClassifier
        sampler = mod.model.model.jem.sampler
    elseif mod.model.model.model isa JointEnergyClassifier
        sampler = mod.model.model.model.jem.sampler
    else
        K = length(counterfactual_data.y_levels)
        input_size = size(selectdim(counterfactual_data.X, ndims(counterfactual_data.X), 1))
        𝒟x = Uniform(extrema(counterfactual_data.X)...)
        𝒟y = Categorical(ones(K) ./ K)
        sampler = ConditionalSampler(𝒟x, 𝒟y; input_size=input_size)
    end
    opt = ImproperSGLD()
    f(x) = logits(mod, x)

    n_iter = 200
    _w = 1500
    plts = []
    neach = 10
    for i in 1:10
        x = sampler(f, opt; niter=n_iter, n_samples=neach, y=i)
        plts_i = []
        for j in 1:size(x, 2)
            xj = x[:,j]
            xj = reshape(xj, (n_digits, n_digits))
            plts_i = [plts_i..., Plots.heatmap(rotl90(xj), axis=nothing, cb=false)]
        end
        plt = Plots.plot(plts_i..., size=(_w,0.10*_w), layout=(1,10))
        plts = [plts..., plt]
    end
    plt = Plots.plot(plts..., size=(_w,_w), layout=(10,1), plot_title=mod_name)
    display(plt)

    # Test performance:
    test_data = load_mnist_test()
    test_data.X = pre_process.(test_data.X)
    f1 = CounterfactualExplanations.Models.model_evaluation(M, test_data)
    println("F1 score (test): $(round(f1,digits=3))")
end
```

>>>>

```{julia}
# Data:
n_obs = 10000
counterfactual_data = load_mnist(n_obs)
counterfactual_data.X = pre_process.(counterfactual_data.X)
X, y = CounterfactualExplanations.DataPreprocessing.unpack_data(counterfactual_data)
X = table(permutedims(X))
labels = counterfactual_data.output_encoder.labels
input_dim, n_obs = size(counterfactual_data.X)
n_digits = Int(sqrt(input_dim))
output_dim = length(unique(labels))
```

@@

```{julia}
# Plot generated samples:

for (mod_name, mod) in model_dict
    if ECCCo._has_sampler(mod)
        sampler = ECCCo._get_sampler(mod)
    else
        K = length(counterfactual_data.y_levels)
        input_size = size(selectdim(counterfactual_data.X, ndims(counterfactual_data.X), 1))
        𝒟x = Uniform(extrema(counterfactual_data.X)...)
        𝒟y = Categorical(ones(K) ./ K)
        sampler = ConditionalSampler(𝒟x, 𝒟y; input_size=input_size)
    end
    opt = ImproperSGLD()
    f(x) = logits(mod, x)

    n_iter = 200
    _w = 1500
    plts = []
    neach = 10
    for i in 1:10
        x = sampler(f, opt; niter=n_iter, n_samples=neach, y=i)
        plts_i = []
        for j in 1:size(x, 2)
            xj = x[:,j]
            xj = reshape(xj, (n_digits, n_digits))
            plts_i = [plts_i..., Plots.heatmap(rotl90(xj), axis=nothing, cb=false)]
        end
        plt = Plots.plot(plts_i..., size=(_w,0.10*_w), layout=(1,10))
        plts = [plts..., plt]
    end
    plt = Plots.plot(plts..., size=(_w,_w), layout=(10,1), plot_title=mod_name)
    savefig(plt, joinpath(output_images_path, "mnist_generated_$(mod_name).png"))
    display(plt)
end
```

```{julia}
# Evaluate models:

measure = Dict(
    :f1score => multiclass_f1score, 
    :acc => accuracy, 
    :precision => multiclass_precision
)
model_performance = DataFrame()
for (mod_name, mod) in model_dict
    # Test performance:
    test_data = load_mnist_test()
    test_data.X = pre_process.(test_data.X)
    _perf = CounterfactualExplanations.Models.model_evaluation(mod, test_data, measure=collect(values(measure)))
    _perf = DataFrame([[p] for p in _perf], collect(keys(measure)))
    _perf.mod_name .= mod_name
    model_performance = vcat(model_performance, _perf)
end
Serialization.serialize(joinpath(output_path,"mnist_model_performance.jls"), model_performance)
CSV.write(joinpath(output_path, "mnist_model_performance.csv"), model_performance)
model_performance
```

```{julia}
Random.seed!(123)

# Set up search:
factual = 8
x = reshape(counterfactual_data.X[:,rand(findall(labels.==factual))],input_dim,1)
target = 3
γ = 0.9
T = 100

# ECCCo:
λ=[0.5,0.1,0.5]
temp=0.1
η=0.01

# Generate counterfactuals using ECCCo generator:
generator = ECCCoGenerator(
    λ=λ, 
    temp=temp, 
    opt=Flux.Optimise.Adam(η),
)

ces = Dict()
for (mod_name, mod) in model_dict
    ce = generate_counterfactual(
        x, target, counterfactual_data, mod, generator; 
        decision_threshold=γ, max_iter=T,
        initialization=:identity,
        converge_when=:generator_conditions,
    )
    ces[mod_name] = ce
end

for (_name,ce) in ces
    x = CounterfactualExplanations.counterfactual(ce)
    _phat = target_probs(ce)
    _title = "$_name (p̂=$(round(_phat[1]; digits=3)))"
    plt = Plots.plot(
        convert2image(MNIST, reshape(x,28,28)),
        axis=nothing, 
        size=(img_height, img_height),
        title=_title
    )
    plts = [plts..., plt]
end
plt = Plots.plot(plts...; size=(img_height*length(plts),img_height), layout=(1,length(plts)))
display(plt)
```
